---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# stackcallr

<!-- badges: start -->
<!-- badges: end -->

`stackcallr` inspects the call tree of an R function without running it.
Point it at a function name and it statically resolves every function
that function calls, the functions those call, and so on, including S3
methods reached through `UseMethod()` dispatch.

## Installation

`stackcallr` isn't on CRAN. Install the development version from GitHub:

``` r
# install.packages("pak")
pak::pak("mjfrigaard/stackcallr")
```

## Usage

`call_stack()` takes the name of a function as a string, bare or
namespaced, and returns the tree of functions it calls. The example
below stops two levels down:

```{r example}
library(stackcallr)

call_stack("stats::sd", max_depth = 2)
```

For a generic, `call_stack()` resolves `UseMethod()` dispatch into the
actual registered methods, rather than stopping at `UseMethod` itself.
Every `mean.*` node below is a real method, ready to be expanded further
with a larger `max_depth`:

```{r generic-example}
call_stack("mean", max_depth = 1)
```

Recursion depth is controlled with `max_depth` (default `Inf`, which
follows every call down to primitives or a detected cycle).

## Shiny module trees

`call_tree_dir()` and `call_tree_app()` build the same kind of tree for a
Shiny app, either from a directory of `.R` files on disk or from a live
`shiny.appobj`. Both take a `modules_only` argument that prunes the tree
down to functions calling `NS()`/`moduleServer()`, splicing out plain
wrapper functions so real modules underneath them stay visible:

```{r modules-example}
app_dir <- tempfile("app")
dir.create(app_dir)
writeLines(
  'launch <- function() { shinyApp(ui = app_ui(), server = app_server) }',
  file.path(app_dir, "launch.R")
)
writeLines('app_ui <- function() { mod_a_ui("a") }', file.path(app_dir, "app_ui.R"))
writeLines(
  'app_server <- function(input, output, session) { mod_a_server("a") }',
  file.path(app_dir, "app_server.R")
)
writeLines('mod_a_ui <- function(id) { ns <- NS(id) }', file.path(app_dir, "mod_a_ui.R"))
writeLines(
  'mod_a_server <- function(id) { moduleServer(id, function(input, output, session) {}) }',
  file.path(app_dir, "mod_a_server.R")
)

call_tree_dir(app_dir, root = "launch", modules_only = TRUE)
```

Neither function requires shiny to be installed for `stackcallr` itself
to work; shiny is in `Suggests`, needed only if you're building a
`shiny.appobj` to pass in.

## Functions

- `call_stack(fn, max_depth = Inf)` — the exported entry point. Resolves
  `fn` to a function, then recursively walks its body via
  `codetools::findGlobals()`, following S3 dispatch where it finds
  `UseMethod()` calls.
- `call_tree_dir(path, root, modules_only = FALSE)` — builds the call
  tree of every function defined in a directory of `.R` files, parsed
  without loading or running any of it.
- `call_tree_app(app, app_fun, ui_fun, server_fun, modules_only = FALSE)`
  — builds the call tree of a running Shiny app by inspecting a
  `shiny.appobj`'s server function and its enclosing environment.
- `print.call_stack()` — renders the tree returned by any of the three,
  in the style of `lobstr::ast()`.

See `vignette("stackcallr")` for how `call_stack()` resolves a function,
and `vignette("shiny-modules")` for `call_tree_dir()` and
`call_tree_app()`.
