---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# stackcallr

<!-- badges: start -->
<!-- badges: end -->

`stackcallr` inspects the call tree of an R function without running it.
Point it at a function name and it statically resolves every function
that function calls, the functions those call, and so on, including S3
methods reached through `UseMethod()` dispatch.

## Installation

`stackcallr` isn't on CRAN. Install the development version from GitHub:

``` r
# install.packages("pak")
pak::pak("mjfrigaard/stackcallr")
```

## Usage

`call_stack()` takes the name of a function as a string, bare or
namespaced, and returns the tree of functions it calls. The example
below stops two levels down:

```{r example}
library(stackcallr)

call_stack("stats::sd", max_depth = 2)
```

For a generic, `call_stack()` resolves `UseMethod()` dispatch into the
actual registered methods, rather than stopping at `UseMethod` itself.
Every `mean.*` node below is a real method, ready to be expanded further
with a larger `max_depth`:

```{r generic-example}
call_stack("mean", max_depth = 1)
```

Recursion depth is controlled with `max_depth` (default `Inf`, which
follows every call down to primitives or a detected cycle).

## Functions

- `call_stack(fn, max_depth = Inf)` — the exported entry point. Resolves
  `fn` to a function, then recursively walks its body via
  `codetools::findGlobals()`, following S3 dispatch where it finds
  `UseMethod()` calls.
- `print.call_stack()` — renders the returned tree, in the style of
  `lobstr::ast()`.

See `vignette("stackcallr")` for how resolution works and what it can't
resolve.
