---
title: "Shiny module trees"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Shiny module trees}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
has_shiny <- requireNamespace("shiny", quietly = TRUE)
```

```{r setup}
library(stackcallr)
```

## Two functions, two entry points

`call_stack()` starts from a function that R can already find by name. A
Shiny app usually isn't in that position: its functions live in a
directory of `.R` files that hasn't been installed, or inside the
environment of an app object that was built but never launched.

`stackcallr` covers both cases with two functions:

- `call_tree_dir()` reads a directory of `.R` files from disk and builds
  the call graph among the functions defined there, without evaluating
  any of it.
- `call_tree_app()` inspects a `shiny.appobj` at runtime and builds the
  call graph among the functions visible to its server function.

Both return an object of class `call_stack`, the same class
`call_stack()` returns, so both print as the same block-character tree.
There's no separate renderer to learn.

## Scanning a directory with `call_tree_dir()`

`call_tree_dir(path, root, modules_only)` parses every `.R` file in
`path`, extracts each top-level function definition
(`name <- function(...) ...`), and builds the call graph among those
locally-defined functions. It returns the tree of calls reachable from
`root`.

Nothing in the scanned files is loaded, sourced, or run. That's what
makes it usable on a package's `R/` directory, or on a Shiny app's
source, without installing it first.

The code below writes a throwaway app into a temporary directory: a
`launch()` function that assembles the app, an `app_ui()`/`app_server()`
pair, and one module, `mod_a_ui()`/`mod_a_server()`.

```{r}
app_dir <- tempfile("app")
dir.create(app_dir)

writeLines(
  'launch <- function() { shinyApp(ui = app_ui(), server = app_server) }',
  file.path(app_dir, "launch.R")
)
writeLines(
  'app_ui <- function() { mod_a_ui("a") }',
  file.path(app_dir, "app_ui.R")
)
writeLines(
  'app_server <- function(input, output, session) { mod_a_server("a") }',
  file.path(app_dir, "app_server.R")
)
writeLines(
  'mod_a_ui <- function(id) { ns <- NS(id) }',
  file.path(app_dir, "mod_a_ui.R")
)
writeLines(
  c(
    'mod_a_server <- function(id) {',
    '  moduleServer(id, function(input, output, session) { helper() })',
    '}'
  ),
  file.path(app_dir, "mod_a_server.R")
)
writeLines(
  'helper <- function() 1',
  file.path(app_dir, "helper.R")
)
```

Rooting the tree at `launch` gives every locally-defined function
reachable from it:

```{r}
call_tree_dir(app_dir, root = "launch")
```

Note that `helper()` shows up under `mod_a_server()`, and that
`shinyApp()` does not show up at all. Only functions defined in the
scanned files become nodes; calls out to `shiny` or any other package
are ignored, since there's no local definition to expand.

If `root` isn't one of the top-level functions found in `path`, the
function errors rather than returning an empty tree.

## Showing only modules with `modules_only`

Both functions take `modules_only`. When `TRUE`, the tree is pruned to
`root` plus the functions that call `shiny::NS()` or
`shiny::moduleServer()`. That pair of names is the static test for "this
function is a Shiny module": it's a pattern match on the parsed
expression, not a call into shiny.

```{r}
call_tree_dir(app_dir, root = "launch", modules_only = TRUE)
```

The interesting part is what happened to `app_ui()` and `app_server()`.
Neither calls `NS()` or `moduleServer()` directly, so neither survives
the prune, but the modules underneath them are still in the tree. When a
node is dropped, its children are spliced up into its parent's position
rather than dropped with it. Without that, a real module reached only
through a plain wrapper function would vanish from the pruned tree, and
`modules_only` would be close to useless on any app with a UI assembly
layer.

`helper()` is gone, because it isn't a module and has no module
descendants to promote.

## Inspecting a live app with `call_tree_app()`

`call_tree_app(app, app_fun, ui_fun, server_fun, modules_only)` works on
a `shiny.appobj`, the object returned by `shiny::shinyApp()` or
`shiny::shinyAppDir()`. Instead of reading files, it extracts the app's
server function, walks the server's enclosing environment chain to
collect every visible function, and builds the call graph among those.

Constructing a `shiny.appobj` does not start a server, so the app below
is safe to build in a knitted document. It defines a module and a
UI/server pair, then passes both to `shiny::shinyApp()`.

```{r eval = has_shiny}
library(shiny)

app <- local({
  mod_a_ui <- function(id) {
    ns <- NS(id)
    tagList(textOutput(ns("out")))
  }
  mod_a_server <- function(id) {
    moduleServer(id, function(input, output, session) {
      output$out <- renderText("hello")
    })
  }
  app_ui <- function() fluidPage(mod_a_ui("a"))
  app_server <- function(input, output, session) mod_a_server("a")

  shinyApp(ui = app_ui, server = app_server)
})

call_tree_app(app)
```

The root node is labelled `"app"` by default. The app object doesn't
store the name of the function that built it, so `app_fun` is display
text only:

```{r eval = has_shiny}
call_tree_app(app, app_fun = "launch", modules_only = TRUE)
```

### Naming the UI function

`ui_fun` and `server_fun` default to `NULL`, which means
`call_tree_app()` recovers the names itself by comparing function bodies
against the functions it collected from the environment.

For the server function this is reliable. For the UI it depends on how
`ui` was passed to `shinyApp()`. Passing a bare function reference
(`shinyApp(ui = app_ui, ...)`) stores the function, and the name can be
recovered. Passing a pre-evaluated call (`shinyApp(ui = app_ui(), ...)`)
stores the resulting tag list, and there's no function left to match, so
the UI branch is dropped.

That second form is common. When you're using it, name the UI function
explicitly:

```{r eval = has_shiny}
app2 <- local({
  mod_a_ui <- function(id) tagList(textOutput(NS(id)("out")))
  mod_a_server <- function(id) {
    moduleServer(id, function(input, output, session) {})
  }
  app_ui <- function() fluidPage(mod_a_ui("a"))
  app_server <- function(input, output, session) mod_a_server("a")

  shinyApp(ui = app_ui(), server = app_server)
})

call_tree_app(app2, ui_fun = "app_ui")
```

## How resolution works

Neither function runs the app. `call_tree_dir()` works entirely from
parsed source:

1. Every `.R` file in `path` is parsed with `parse()`. Files that fail
   to parse are skipped rather than raising an error.
2. Each top-level expression is checked for a function assignment
   (`name <- function(...)`, `name = function(...)`, or
   `assign("name", function(...))`), and the name and unevaluated body
   are kept.
3. Each body is walked recursively for references to the other collected
   names, counting both direct calls (`fun()`) and bare references
   (`fun`), so a function passed as an argument still registers as an
   edge.
4. The tree is built depth-first from `root`, skipping any name already
   on the current path, so recursive or mutually recursive functions
   don't loop forever.
5. When `modules_only = TRUE`, each body is tested for `NS()` or
   `moduleServer()` (bare or namespaced), and non-matching nodes are
   spliced out.

`call_tree_app()` follows the same steps 3 through 5. It differs only in
where the functions come from: `app$serverFuncSource()` gives the server
function, `environment()` gives its enclosure, and the environment chain
is walked upward, collecting functions until it reaches `globalenv()`,
`baseenv()`, or a loaded package namespace.

## What it can't inspect

`call_tree_app()` requires a `shiny.appobj`. It cannot inspect the
`ShinyAppHandle` returned by `shiny::startApp()` (added in shiny
1.14.0). That handle is a runtime control object, exposing `stop()`,
`status()`, `url()`, and `result()` for an app that may be running in a
separate process, and it holds no reference back to the app's UI,
server, or environment. Passing one raises an error that says so, and
tells you to pass the `shiny.appobj` itself, before calling `startApp()`
on it.

Two other limits are worth knowing:

- `call_tree_dir()` only sees top-level function definitions. A function
  defined inside another function, or built by a factory, isn't a node.
- `call_tree_app()` only sees what's visible from the server's
  environment chain. Functions that live in an unrelated environment,
  or in a package namespace, stop the walk rather than being expanded.

## Shiny isn't a dependency

Neither function requires shiny to be installed for `stackcallr` to
work. `call_tree_dir()` matches on the names `NS` and `moduleServer` in
parsed code, which is text, not a call. `call_tree_app()` identifies the
object it's handed with `inherits()` and reaches into it with `$`; there
is no `shiny::` call anywhere in its body.

Shiny is in `Suggests`, and you need it only if you're building a
`shiny.appobj` to pass in.

## Learning more

See `?call_tree_dir` and `?call_tree_app` for the full argument
reference, and `vignette("stackcallr")` for `call_stack()`, which covers
the case these two complement: a single function R can already resolve
by name.
