---
title: "Introduction to stackcallr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to stackcallr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(stackcallr)
```

## Why stackcallr

When you're reading unfamiliar code, one of the first questions is: what
does this function actually call? For a simple function you can just read
the body. For an S3 generic like `dplyr::select()`, the body is a single
line, `UseMethod("select")`, and the real work happens in a method you'd
have to go find yourself.

`stackcallr` answers that question directly. Give `call_stack()` the name
of a function, and it returns the tree of every function it calls, the
functions those call, and so on, resolving S3 dispatch along the way.

## The basic call

`call_stack()` takes a single string naming a function, either bare
(`"mean"`) or namespaced (`"pkg::fun"`, `"pkg:::internal_fun"`):

```{r}
call_stack("mean.default", max_depth = 1)
```

The result is a `call_stack` object: a tree where each node has a `name`
and a list of `calls` (its children). Printing it draws the tree.

## Limiting depth

Some call trees are deep. `max_depth` caps how many levels are followed,
so `max_depth = 1` returns only the functions called directly by `sd()`:

```{r}
call_stack("stats::sd", max_depth = 1)
```

Raising it to `2` expands each of those nodes one level further:

```{r}
call_stack("stats::sd", max_depth = 2)
```

The default, `max_depth = Inf`, follows every call all the way down,
stopping only at primitives (functions with no R-level body to inspect)
and at calls it has already followed once, so recursive or mutually
recursive functions don't loop forever.

## Following S3 dispatch

Static analysis alone can't tell you what `UseMethod("select")` actually
runs, since that depends on the class of the argument at call time.
`call_stack()` handles this case specifically: when a function's body
calls `UseMethod()`, it looks up every method registered for that
generic and follows each one, in place of a single unhelpful `UseMethod`
leaf.

```{r}
call_stack("mean", max_depth = 1)
```

Each `mean.*` entry above is a real method, resolved and ready to be
expanded further with a larger `max_depth`.

## How resolution works

`call_stack()` never runs the function you point it at. It works
entirely from source:

1. The named function is resolved to a closure via `get()` /
   `getExportedValue()`.
2. Its body is parsed with `codetools::findGlobals()` to list the
   functions it references, with R's own syntax (`{`, `if`, `<-`, `+`,
   ...) filtered out, since those aren't calls in any meaningful sense.
3. Each referenced name is looked up in the function's own lexical
   scope, the same place R itself would resolve it, so calls to
   internal, unexported helpers are found correctly.
4. If the body calls `UseMethod()`, the generic's registered methods are
   found and substituted in for that call.
5. The process repeats for every function found, down to `max_depth` or
   until a cycle is detected.

## What it can't resolve

`UseMethod()` is the one form of dynamic dispatch `call_stack()` follows.
Everything else that decides which function to run at call time stays
unresolved, and shows up as a leaf rather than being expanded:

- `NextMethod()` calls, since the next method depends on the class vector
  in play at the time.
- S4 and R6 dispatch.
- Functions named only as data, such as those passed to `do.call()`,
  `match.fun()`, or `Recall()`.

A leaf in the printed tree, then, means one of three things: a primitive,
a cycle already followed once, or a call that static analysis can't see
through.

## Learning more

See `?call_stack` for the full argument reference, and `?print.call_stack`
for how the tree is drawn.
